---
title: "aneur: generalized additive mixture model (GAMM)"
output: 
  pdf_document: 
    latex_engine: xelatex
    keep_tex: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aortic Aneurysm Progression Data

This dataset contains longitudinal measurements of grades of aortic aneurysms, measured by ultrasound examination of the diameter of the aorta.

A data frame containing 4337 rows, with each row corresponding to an ultrasound scan from one of 838 men over 65 years of age.


* ptnum	(numeric)	Patient identification number   

* age	(numeric)	Recipient age at examination (years)  

* diam	(numeric)	Aortic diameter   

* state	(numeric)	State of aneurysm.   


The states represent successive degrees of aneurysm severity, as indicated by the aortic diameter.

* State 1	Aneurysm-free	< 30 cm  

* State 2	Mild aneurysm	30-44 cm  

* State 3	Moderate aneurysm	45-54 cm  

* State 4	Severe aneurysm	> 55 cm  


683 of these men were aneurysm-free at age 65 and were re-screened every two years. The remaining men were aneurysmal at entry and had successive screens with frequency depending on the state of the aneurysm. Severe aneurysms are repaired by surgery.



```{R, eval=TRUE, fig.show="hold", echo=FALSE, include=TRUE, out.width="100%", warning=FALSE, message=FALSE}
library(rstan) ### rstan
library(dplyr)
library(ggplot2) ### plots
library(splines) ### splines
library(splines2) ### splines
library(msm) ### aneur data
library(loo)  ### compute information criteria
```

```{R, eval=TRUE, fig.show="hold", echo=FALSE, include=TRUE, out.width="100%", warning=FALSE, message=FALSE}
setwd("/Users/lizbethna/Documents/Sabatico/Examples/Aneur/")
```

```{r, echo=TRUE, eval=TRUE, fig.show="hold", include=TRUE, out.width="100%", warning=FALSE, message=FALSE} 
data(aneur)
attach(aneur)
head(aneur)
tail(aneur)
#help(aneur)
dim(aneur)

(N = n_distinct(aneur$ptnum))   # subjects
(K = max(table(aneur$ptnum)))   # times
table(table(aneur$ptnum))

J = 4   # categories
### data having width format representation
Y_diam = array(NA,dim=c(N,K)) 
Y_state = array(NA,dim=c(N,K))
X_age = array(NA,dim=c(N,K)) 
Ki = table(aneur$ptnum)
Ni = c(0,cumsum(Ki))+1
for(i in 1:N){
	aneur_i = aneur[aneur$ptnum==i,]
	for(k in 1:Ki[i]){
		Y_diam[i,k] = aneur_i$diam[k] 
		Y_state[i,k] = aneur_i$state[k]
		X_age[i,k] = aneur_i$age[k]   
	}  
} 
```

```{r, echo=TRUE, eval=TRUE, fig.show="hold", include=TRUE, out.width="100%", warning=FALSE, message=FALSE} 
### see some data having width format representation
(Y_diam[16:18,1:8])
(Y_state[16:18,1:8]) 
(X_age[16:18,1:8]) 
(Ki[16:18]) 
```


```{r eval=TRUE, echo=TRUE, fig.show="hold", message=FALSE, warning=FALSE, include=TRUE, out.width="100%"}
### Considering only data having more than one screen (diam!=29, or diam<29 & dim>29)
idx3 = c()
for(i in 1:N){
  if( min(Y_diam[i,1:Ki[i]])!=max(Y_diam[i,1:Ki[i]])){
    idx3 = c(idx3,i) 
  }
}
Y3_diam = Y_diam[idx3,]
Y3_state = Y_state[idx3,] 
X3_age = X_age[idx3,]
N3 = length(idx3)
Ki3 = Ki[idx3]
### data used for the analysis 
aneur3 = aneur%>%filter(aneur$ptnum%in%idx3)
```

```{r eval=TRUE, echo=TRUE, fig.show="hold", message=FALSE, warning=FALSE, include=TRUE, out.width="100%"} 
### plot some subjects
Ki3[c(81,96,137)]
```

```{r, echo=FALSE, include=TRUE, evaluate=TRUE} 
plot(0, type="n", 
     xlim=c(min(X3_age[,],na.rm=TRUE),max(X3_age[,],na.rm=TRUE)), 
     ylim=c(min(Y3_diam[,],na.rm=TRUE),max(Y3_diam[,],na.rm=TRUE)), 
     xlab="Age at examination",ylab="Aortic diameter (mm)")
#plot(0, xlim=c(60,79.5), ylim=c(10,63), type="n", xlab="Age at examination",ylab="Aortic diameter (mm)")

abline(h=c(30,45,55),lty=2,lwd=2)
text(61.5,25,"Stage 1")
text(61.5,37,"Stage 2")
text(61.5,50,"Stage 3")
text(61.5,59,"Stage 4")

points(X3_age[81,],Y3_diam[81,], pch=19,cex=0.8,col="red")
lines(X3_age[81,],Y3_diam[81,], lwd=1.5,col="red")
points(X3_age[96,],Y3_diam[96,], pch=17,cex=0.8,col="magenta")
lines(X3_age[96,],Y3_diam[96,], lwd=1.5,col="magenta")
points(X3_age[137,],Y3_diam[137,], pch=15,cex=0.8,col="blue")
lines(X3_age[137,],Y3_diam[137,], lwd=1.5,col="blue")

legend("bottom",c("Subject 690","Subject 705","Subject 746"),
       pch=c(19,17,15),col=c("red","magenta","blue"),lty=1,ncol=3,cex=0.8) 
```


```{r eval=TRUE, echo=TRUE, fig.show="hold", message=FALSE, warning=FALSE, include=TRUE, out.width="100%"} 
ggplot(data=aneur3, mapping=aes(x=age,y=diam,group=ptnum)) + 
  geom_line(color=aneur3$ptnum) + theme_bw() +
  xlab("Age at examination in years") + ylab("Aortic diameter in mm") + 
  ggtitle("Profiles aortic diameter by patient")  
ggplot(data=aneur3, mapping=aes(x=age,y=state,group=ptnum)) + 
  geom_line(color=aneur3$ptnum) + theme_bw() +
  xlab("Age at examination in years") + ylab("States of aneurysm severity") + 
  ggtitle("Profiles states of aneurysm severity by patient")  
```


\clearpage

# Generalized additive mixture model (GAMM) 

It is possible to consider a continuous response variable (``diam'') or an ordinal response variable (``state''), where the explanatory variable is continuous (``age''). Then, the GAMM models are the following. 
\vspace{1cm}

The GAMM for the continuous response variable $diam_{it}$, with random slope:
\begin{eqnarray*}
diam_{it} &=& \beta_0 + f_1(age_{it}) +age_{it} \times b_{1i} + \varepsilon_{it}, 
\\ && 
b_{1i}\sim\mathrm{N}(0,\psi^2), 
\qquad 
\varepsilon_{it} \sim\mathrm{N}(0,\sigma^2)  ,
\end{eqnarray*}
where $\beta_0$ is the intercept and $f_1$ is the smoothing function for the common fixed effects; 
$b_{1i}$ is the random slope to consider that subjects have different growth rates;
$\psi^2$ is the variance for the random slope, and $\sigma^2$ is the variance for the errors. 
Note that random intercepts are not needed to model the data, 
and observations for the same subject are independent, i.e., 
$\varepsilon_{it_1}$ is independent of $\varepsilon_{it_2}$ for $\varepsilon_{it_1}\neq\varepsilon_{it_2}$.
\vspace{1cm}


The ordinal response $state_{it}$ is modelled in terms of the cumulative probabilities $P(state_{it} \leq j | \boldsymbol{b}_i)$ by using the proportional odds model,  
 \begin{eqnarray*}
 P(state_{it} \leq j | \boldsymbol{b}_i) &=& \eta_{it,j}, 
 \end{eqnarray*}
 with $j=1,2,3$, subject to 
\begin{eqnarray*}
\eta_{it,j} &=& \kappa_{j} + f_1(age_{it}) +age_{it} \times b_{1i} , 
\qquad 
b_{1i}\sim\mathrm{N}(0,\psi^2), 
\end{eqnarray*}
where the constraints are such that $f_1$ is a non-decreasing smoothing function and $b_{1i}>0$, and for the breakpoints $\kappa_{j}<\kappa_{j+1}$. 


```{R, eval=TRUE, fig.show="hold", echo=TRUE, include=TRUE, out.width="100%", warning=FALSE, message=FALSE}
### center variables
y = aneur3$diam -29  
x1 = aneur3$age -60 
x2 = aneur3$age -60 
id = as.numeric(as.factor(aneur3$ptnum))

n = length(y)
N = n_distinct(id)
Ni = c(0,cumsum(table(id)))+1
k1 = 4 
k2 = 4 
knots1 = quantile(x1, c(0.33,0.67))
knots2 = quantile(x2, c(0.33,0.67))
```



# Generate the design matrix $X$ for the penzalized B-splines 

Note that the smoothing function $f(x)$ is represented as: 
\begin{eqnarray*}
f(x) 
&=& \sum_{j=1}^{h_1} \beta_{1j}\boldsymbol{I}_{1j}(x)   
\end{eqnarray*} 
for $\beta_{1j}$  unknown parameters.

The number of knots $K$ is chosen a priori. 

```{R, eval=TRUE, fig.show="hold", echo=TRUE, include=TRUE, out.width="100%", warning=FALSE, message=FALSE}
# Generate a basis matrix for Natural Cubic Splines 
X2 <- ns(x = x2, knots = knots2, intercept = TRUE) 
###X2 = (X2-mean(X2))/sd(X2) 
matplot(x2, X2) 
```


# Generate the design matrix $XI$ for the penzalized I-splines 

Note that the smoothing function $f(x)$ is represented as: 
```{=tex}
\begin{eqnarray*}
f_1(x_1)&&\sum_{j=1}^{h_1} \beta_{1j}\boldsymbol{I}_{1j}(x_1)  \\
\boldsymbol{I}_{1j}(x_1) &=& \int_{x_0}^{x_1} \boldsymbol{B}_{1j}(u) d_{u}
\end{eqnarray*}
```


```{R, eval=TRUE, fig.show="hold", echo=TRUE, include=TRUE, out.width="100%", warning=FALSE, message=FALSE}
### ibs: integrated basis splines 
### degree = 3 cubic splines
XI1 <- ibs(x1, knots = knots1, degree = 1, intercept = TRUE) 
###XI1 = (XI1-mean(XI1))/sd(XI1) 
matplot(x1, XI1)
abline(v = knots1, h = knots1, lty = 2, col = "gray") 
```


# Define the penalizations $S1$ and $S2$

The flexibility of $f$ is controlled by  $K$, from a quadratic penalization as:  
\begin{eqnarray*}
\sum_{j} \lambda_j \boldsymbol{\beta}^{T} \boldsymbol{S}_j \boldsymbol{\beta} 
\end{eqnarray*}
where the $\boldsymbol{S}_j$ are matrix with known coefficients, and parameters $\lambda_j$ are smoothing parameters that should be estimated. 

```{R, eval=TRUE, fig.show="hold", echo=TRUE, include=TRUE, out.width="50%", warning=FALSE, message=FALSE}
#No es el óptimo, pero funciona.  
#“k” number of b-splines 
#“d” order of difference 

# Produce the matrix of differences:   
diffMatrix = function(k, d = 2){
  if( (d<1) || (d %% 1 != 0) )stop("d must be a positive integer value"); 
  if( (k<1) || (k %% 1 != 0) )stop("k must be a positive integer value"); 
  if(d >= k)stop("d must be lower than k"); 
  out = diag(k); 
  for(i in 1:d){ 
    out = diff(out); 
  } 
  return(out) 
} 
(D1 = diffMatrix(k=k1, d=2))
(D2 = diffMatrix(k=k2, d=2))
### Matrix of penalization
(S1 = t(D1)%*%D1 + diag(1,k1)*10e-4) 
(S2 = t(D2)%*%D2 + diag(1,k2)*10e-4) 
```

 
\clearpage

# GAMM with monotone constrains  

## Data, inits and parameters  

```{R, eval=TRUE, fig.show="hold", echo=TRUE, include=TRUE, out.width="100%", warning=FALSE, message=FALSE}
data.lme.add.incr <- list( y = y , id = id ,
               n = length(y) , N = N , Ni = Ni, k1=k1, 
              XI1 = XI1, x1 = x1, 
              zero = rep(0,1+k1),  S1 = S1  )  
inits.lme.add.incr <- function(){	list( 
  "b1" = abs(rnorm(k1,0,0.1)),  
  "invtau2" = rgamma(1,1,1) ,
  "lambda" = rgamma(1,1,1) ,
  "invsig2" = rgamma(1,1,1) 
)	} 
param.lme.add = c("b0","b1", "invtau2","tau2","tau", "lambda","rho", "invsig2","sig2","sigma")
``` 

## Fit the model

```{R, eval=TRUE, fig.show="hold", echo=TRUE, include=TRUE, out.width="100%", warning=FALSE, message=FALSE} 
fit.lme.add.incr.reslope <- stan("gamm_aneur_lme_add_incr_reslope.stan",
            data=data.lme.add.incr,
            chains=3,warmup=300,iter=600,thin=2,cores=4,
            init= inits.lme.add.incr) 
```

## Results 

```{R, eval=TRUE, fig.show="hold", echo=TRUE, include=TRUE, out.width="100%", warning=FALSE, message=FALSE}
print(fit.lme.add.incr.reslope, pars=param.lme.add) 
```

## Plots

```{R, eval=TRUE, fig.show="hold", echo=TRUE, include=TRUE, out.width="100%", warning=FALSE, message=FALSE}
stan_trace(fit.lme.add.incr.reslope, pars=param.lme.add)
```

```{R, eval=TRUE, fig.show="hold", echo=TRUE, include=TRUE, out.width="50%", warning=FALSE, message=FALSE}
stan_dens(fit.lme.add.incr.reslope, pars=c("b0"))
stan_dens(fit.lme.add.incr.reslope, pars=c("b1"))
```

```{R, eval=TRUE, fig.show="hold", echo=TRUE, include=TRUE, out.width="50%", warning=FALSE, message=FALSE}
stan_plot(fit.lme.add.incr.reslope, point_est = "mean", show_density = TRUE,
          pars=c("invtau2","tau2","tau") )
stan_plot(fit.lme.add.incr.reslope, point_est = "mean", show_density = TRUE,
          pars=c("invsig2","sig2","sigma") )
stan_plot(fit.lme.add.incr.reslope, point_est = "mean", show_density = TRUE,
          pars=c( "lambda") )
stan_plot(fit.lme.add.incr.reslope, point_est = "mean", show_density = TRUE,
          pars=c("rho") )
```


```{R, eval=TRUE, fig.show="hold", echo=TRUE, include=TRUE, out.width="100%", warning=FALSE, message=FALSE}
pairs(fit.lme.add.incr.reslope, pars = c("b0","b1"), las = 1)
``` 


```{R, eval=TRUE, fig.show="hold", echo=TRUE, include=TRUE, out.width="100%", warning=FALSE, message=FALSE}
mu1 = get_posterior_mean(fit.lme.add.incr.reslope,"mu1")
aneur3$mu1_all = 29 + mu1[,"mean-all chains"]

ggplot(data=aneur3, 
       mapping=aes(x=age,y=mu1_all,group=ptnum)) + 
  geom_point() +
  theme_bw() +
  xlab("Age at examination in years") + ylab("Aortic diameter in mm") + 
  ggtitle("Estimated mean aortic diameter")  
```


```{R, eval=TRUE, fig.show="hold", echo=TRUE, include=TRUE, out.width="100%", warning=FALSE, message=FALSE}
mu2 = get_posterior_mean(fit.lme.add.incr.reslope,"mu2")
aneur3$mu2_all = 29 + mu2[,"mean-all chains"]

ggplot(data=aneur3, 
       mapping=aes(x=age,y=mu2_all,group=ptnum)) + 
  geom_line(color=aneur3$ptnum) +
  theme_bw() +
  xlab("Age at examination in years") + ylab("Aortic diameter in mm") + 
  ggtitle("Estimated profiles aortic diameter by patient")  
```


## Information criteria

```{R, eval=TRUE, fig.show="hold", echo=TRUE, include=TRUE, out.width="100%", warning=TRUE, message=TRUE} 
### fitted model
loo_sample_lme = fit.lme.add.incr.reslope
### we have to extract those log-likelihood terms that we so carefully had Stan calculate for us:
log_lik_lme =extract_log_lik(loo_sample_lme, merge_chains = F)
r_eff_lme =relative_eff(log_lik_lme)
###  look at the results for each model, first the one with mu estimated:
(loo_lme <- loo(log_lik_lme, r_eff=r_eff_lme))
```



